Automatic Matching of Bullet Lands

### 2017-02-02 arxiv/imaging-paper.tex
Remove line numbers
One more linenumber removal
Arxiv update and AoAS resubmission

\linenumbers
%\tableofcontents
%\newpage

@@

The purpose of our paper is to present an automatic matching routine that allows for a completely objective assessment of the quality of a match. Our algorithm is fully open source and available on GitHub \citep{x3prplus}. This transparency allows for a greater understanding of the individual steps involved in the bullet matching process, and allows other forensic examiners, as well as outside observers, examine the factors that separate known bullet matches from non-matches. We have chosen to perform the matching on a land-to-land level, rather than bullet-to-bullet. Although doing so introduces an implicit assumption of independence between lands, assuming independence only serves to make the task more challenging.
The original paper on the complete Hamby study already reports the successful use of several computer-assisted methods; aside from a zero false positive rate, no further details on a false-negative error rate for bullets are given nor are error rates for land-to-land matches mentioned. 
% Hamby study: automated systems were successfully used for matching:
%  Intelligent Automation's SciClops - Dr. Ben Bachrach (Maryland, United States);
% Automated Land Identification System (ALIS) - Mr. Tsuneo Uchiyama (Tokyo, Japan);
% Integrated Ballistics Identification System (IBIS) – Mr. Robert Thompson (California, United States);
% BulletTRAX-3DTM - Forensic Technology Scientists (Montreal, Canada) (2 sets)

A procedure for bullet matching using the BulletTrax3D system is described in \citet{roberge:2006}. This study is based on another set of ten consecutively rifled barrels; matches are identified based on a bullet-to-bullet correlation score. The authors state that this process `could be automated', but no implementation of the algorithm is available. Modern automated techniques using 3D images have also been conducted by \citet{riva:2014}, but this work applies to striae found on cartridge cases and not bullets. Other work on 3D images has been conducted by \citet{petraco:2012}, which also describes work carried out on cartridge cases, as well as screwdriver striation patterns, with promising results. We aim to use similar ideas of automated 3D image analysis, and apply this work to bullet surfaces with the modern tools enabled by the R statistical software.

The remainder of this paper is structured as follows: We first discuss two methods of modeling the class structure of the bullet surfaces. %We then describe a web application for visualizing and working with bullet surfaces. 
We then proceed to describing an automatic matching routine which we evaluate on the bullets made available through the Hamby study.

@@

\subsection{Horizontal alignment}
Signatures of each of the two lands, 1-5 and 2-1,  in Figure~\ref{fig:manualmatch-rgl} are shown in Figure~\ref{fig:cross100} extracted at a height of $x = 100\mu m$. Striation marks show up in these representations as peaks and valleys.  The individual characteristics are quite prominent and, again, suggest a match between the lands. A horizontal shift of one of the signatures (result shown in Fig~\ref{fig:crosscutX}) emphasizes the strong similarities. 
%
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Raw bullet land signatures.\label{fig:crosscut}}{%
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\textwidth]{crosscuts-1}

@@

\section{Evaluation}
In order to get a better understanding of  how  the matching algorithm works in known matches and non-matches, we investigate its performance in a test scenario based on the James Hamby study.
As a first step, we automatically assess the quality of each of the lands by  checking that we can identify a stable region on each land. For this, we compute the cross-correlation of signatures extracted from heights 25$\mu m$ apart. For a stable region, we require a minimum of 0.95 for the cross correlation. Four lands from different bullets are flagged as problematic in this respect. A visual inspection (see Figure~\ref{fig:fourflags}) shows that each one of these lands has scratch marks across the surface, also known as `tank rash'.
%
\begin{figure}
  \centering
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 6 Bullet 2-1}
%' <<echo=FALSE>>=
%' b1 <- read.x3p(file.path(datadir, "Br6 Bullet 2-1.x3p"))
%' image(b1$surface.matrix)
%' @
\includegraphics[width=\textwidth]{br6-2-1-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 9 Bullet 2-4}
\includegraphics[width=\textwidth]{br9-2-4-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet B-2}
\includegraphics[width=\textwidth]{b-2-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet Q-4}
\includegraphics[width=\textwidth]{q-4-grey.png}
\end{subfigure}
\caption{\label{fig:fourflags}Images of the four lands that got flagged during the quality assessment. All of them show scratch marks (tank rash) across the striation marks from the barrel. They are excluded from the remainder of the analysis.}
\end{figure}
%
We exclude these four lands from further matching considerations and
 run all remaining lands from the unknown bullets against all remaining lands from known bullets for matches, i.e. we are comparing $15 \times 6 -2 = 90 - 2 = 88$ lands from unknown bullets against $2 \times 10 \times 6 -2 = 120 - 2 = 118$ lands from known bullets, yielding a total of $10,384$ land-to-land comparisons. Out of these comparisons, there are 172 known matches (KM), while the rest are known non-matches (KNM). 
 When things go well, they look like the results in Figure~\ref{fig:hamby-perfect}: Figure~\ref{fig:hamby-perfect}a shows the distribution of the number of maximum consecutive matching striae between land C-3 and all 118 lands from known bullets. Two lands show a high CMS. These correspond to the known matches with C-3, shown in Figures~\ref{fig:hamby-perfect}b and~\ref{fig:hamby-perfect}c. 
 Unfortunately, not all results are as clear cut. 
It might not be reasonable to assume that we can match all lands, but the idea is to try to maximize the number of matches to get an overview of what we might be able to expect from an automated match.

@@

Besides resulting in a probabilistic quantification of matches, random forests also provide an assessment of the importance of each of the features derived from the bullets' 3D topological surface measurements. Figure~\ref{fig:importance} shows an overview of the importance of each variable measured as the mean decrease in the Gini index when the variable in question is included in a tree (for the exact values please refer to Supplement section~\ref{supp:randomforest}). 
\begin{figure}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=.7\textwidth]{unnamed-chunk-1-1}

@@

%The rows in the table are ordered according to descending importance. 
The variables with the most predictive power are cross-correlation and the overall number of matches, followed by the total depth of joint striations $S$ and total number of non-matches. CMS is found only in sixth place.

Besides including results from known matches against known non-matches, we can increase the number of comparisons in the Hamby study to include all possible land-to-land comparisons. This effectively doubles the number of data points available. Comparisons not previously included in fitting the random forest can also be used as an additional source for assessing error rates.  Results for this and a more detailed discussion can be found in Supplement section~\ref{supp:extended}.

\section{Conclusion}

We have presented an algorithm which detects the most prominent but least relevant structure of a bullet from a firearms identification perspective, removes these features, and produces residuals which allow for the easy identification of markings. We have generalized this algorithm to align the residuals from two bullets to automatically determine whether they are matches. We created a random forest model using many of the features to provide a probabilistic assessment of the quality of a match, along with the most relevant features. Matching bullets is clearly not a single step, but rather a sequence of a number of data analysis tasks which each deserve attention. As there is no scientific standard in place at this point in time, our intent is to explain one approach to addressing these tasks, while documenting all steps and providing all code so other researchers and forensic scientists can reproduce and expand on our findings.

The matching algorithm is sensitive to the parameter choices made. The heights at which signatures are extracted (currently 25$\mu m$ apart) to evaluate stability, as well as the cross-correlation factor (currently 0.95) we set as a minimum threshold do affect the final outcome. Another parameter that must be selected is the amount of smoothing when identifying peaks and grooves (currently, a window of 23.4375$\mu m$ is used, corresponding to a window of 7 values to the left and the right of an observation). While we tried to lay out in the paper the impact that each of the parameter choices has on the matching performance, we are still far from an optimized scenario.

The Hamby study serves as our evaluation `database'. It consists of only  35 bullets -- this is obviously not a particularly realistic scenario for an automatic matching procedure, but for now we are unaware of other databases containing bullets in the x3p format.

The feasibility of creating a database of ballistic images that could be used to identify guns used in crimes was evaluated in a 2008 report~\citep{nap:2008} by the National Research Council. The evaluation was investigating the scalability of NIBIN (National Integrated Ballistic Information Network), which uses proprietary matching algorithms provided by IBIS. The bottom line of the report was that in spite of the many technical and practical hurdles, solutions to all but one problem could be found. The problem that remained is that statistically, the quality of the matching algorithm (in this case, of breech-face marks and firing pin impressions) could not withstand a hugely increased number of records while still maintaining a reasonable workload for forensic examiners, who have to examine possible matches suggested by the system. 
The findings of the NRC report on ballistic imaging are based on two-dimensional greyscale images, which the committee argued were not reliable enough for distinguishing between fine marks. This finding coincides with the assessment by \citet{dekinder:2004} based on the IBIS Heritage system. A further re-assessment by \citet{deceuster:2015} came to the same conclusions based on the EvoFinder system. 
The NRC report also found that results from 2D images can be improved when matches are based on 3D images. This is consistent with the importance of features found here: out of the top five features (see Figure~\ref{fig:importance}), only the total number of matches and mismatches are available for a match based on 2D features.

By suggesting an automated algorithm that first removes class characteristics, such as the grooves and the curvature of the bullet to reveal the region of the  land, then identifies peaks and valleys on this land, we reduce subjectivity and with it possible sources of bias. In particular, `the concept of counting striations is subjective and based on experience'~\citep{miller:1998}. The steps outlined in this paper could also help explore other important forensic science problems. In particular, more general toolmark examination can benefit from this detection and removal of the overall structure and the extraction of features from deviations of this structure.

For a fair assessment of the performance of an algorithm, we need transparency. Our matching  algorithm is open: the code is readily available in form of the R package x3prplus~\citep{x3prplus}, and the code to produce this paper is available at \url{http://www.github.com/erichare/imaging-paper}. To understand whether an automated approach along the lines of the one we propose can accurately identify sets of bullets with undistinguishable markings, it will be necessary to assemble a much larger database that includes a wide range of ammunition types, degrees of damage, gun makes, etc. We are unaware of the existence of any such database. In addition to serving as a realistic testbed for the performance of the automated matching algorithm, such a database would also permit testing the underlying, as of yet untested, assumptions of uniqueness and reproducibility of the markings left by a gun on bullets.

>>>>

%\tableofcontents
%\newpage

@@

The purpose of our paper is to present an automatic matching routine that allows for a completely objective assessment of the strength of a match between two bullet lands. While we assess the performance of the algorithm in terms of a binary decision of match vs. non-match using a 50\% probability cut-off, our primary goal is to highlight the features that are statistically associated with matches and non-matches, and to provide a quantitative assessment of this association. In a real-world application of our algorithm, the raw scores would need further analysis and scrutiny, and it is likely a 50\% cut-off would be an inappropriate choice on the basis of reasonable doubt.

Our algorithm is fully open source and available on GitHub \citep{bulletr}. This transparency allows for a greater understanding of the individual steps involved in the bullet matching process, and allows other forensic examiners, as well as outside observers, to examine the factors that discriminate between known bullet matches and non-matches. We have chosen to perform the matching on a land-to-land level, rather than bullet-to-bullet level. Although doing so introduces an implicit assumption of independence between lands, assuming independence only serves to make the task more challenging.

The remainder of this paper is structured as follows: We first briefly review some earlier work. We then discuss two methods of modeling the class structure of the bullet surfaces. Finally, we proceed to describing an automatic matching routine which we evaluate on the bullets made available through the Hamby study.

\section{Previous work}

There have been attempts to develop automatic or semi-automatic matching protocols, but most have focused on breech face and firing pin marks \citep[e.g.][]{riva:2014} or discuss a single attribute for comparison \citep[e.g.][]{vorburger:2011, chu:2011}. Still others refer to proprietary algorithms \citep{roberge:2006}. We briefly review some of this earlier work in what follows.

The original paper on the complete Hamby study already reports the successful use of several computer-assisted methods. However, aside from a zero false positive rate, false-negative error rates for bullets are not given nor are error rates for land-to-land matches mentioned.

\citet{lock2013significance} proposed an approach to quantify similarity of toolmarks. Their algorithm determines an optimal matching window between two toolmark signatures, and then performs a set of both coordinated and independent shifts. Given a match, the coordinated shifts would be expected to yield correlation values higher than those obtained from independent shifts. This is assessed using a Mann-Whitney U Statistic.

A procedure for bullet matching using the BulletTrax3D system is described in \citet{roberge:2006}. Their study used a different set of ten consecutively rifled barrels; matches are identified based on a bullet-to-bullet correlation score. The authors state that this process `could be automated', but no implementation of the algorithm is available.

Modern automated techniques using 3D images have also been proposed by \citet{riva:2014}. However, the authors focused on cartridge cases and not bullets. This might seem like a trivial distinction, but it has implications for the development of the algorithm. The algorithm performs alignment of striae by rotation of the XY plane, which is not generalizable to bullets in which the XY plane is not flat.

Other work on 3D images has been described by \citet{petraco:2012}, who also focus on cartridge cases, as well as screwdriver striation patterns, and by others \citep[e.g.,][]{chu:2010, chu:2011, vorburger:2011}.

@@

\subsection{Horizontal alignment}
Signatures of each of the two lands, 1-5 and 2-1,  in Figure~\ref{fig:manualmatch-rgl} are shown in Figure~\ref{fig:cross100} extracted at a height of $x = 100\mu m$. Striation marks show up in these representations as peaks and valleys.  The individual characteristics are prominent and, again, suggest a match between the lands. A horizontal shift of one of the signatures (result shown in Fig~\ref{fig:crosscutX}) emphasizes the strong similarities between signatures. 
%
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Raw bullet land signatures.\label{fig:crosscut}}{%
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\textwidth]{crosscuts-1}

@@

\section{Evaluation}
In order to get a better understanding of  how  the matching algorithm works in known matches and non-matches, we investigate its performance using the James Hamby study data. As a first step, we automatically assess the quality of each of the lands by  checking that we can identify a stable region on each land. For this, we compute the cross-correlation of signatures extracted from heights 25$\mu m$ apart. For a stable region, we require a minimum of 0.95 for the cross correlation. Four lands from different bullets are flagged as problematic in this respect. A visual inspection (see Figure~\ref{fig:fourflags}) shows that each one of these lands has scratch marks across the surface, also known as `tank rash'.
%
\begin{figure}
  \centering
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 6 Bullet 2-1}
%' <<echo=FALSE>>=
%' b1 <- read_x3p(file.path(datadir, "Br6 Bullet 2-1.x3p"))
%' image(b1$surface.matrix)
%' @
\includegraphics[width=\textwidth]{br6-2-1-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 9 Bullet 2-4}
\includegraphics[width=\textwidth]{br9-2-4-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet B-2}
\includegraphics[width=\textwidth]{b-2-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet Q-4}
\includegraphics[width=\textwidth]{q-4-grey.png}
\end{subfigure}
\caption{\label{fig:fourflags}Images of the four lands that got flagged during the quality assessment. All of them show scratch marks (tank rash) across the striation marks from the barrel. They are excluded from the remainder of the analysis.}
\end{figure}
%
We exclude these four lands from further matching considerations and
 run all remaining lands from the unknown bullets against all remaining lands from known bullets for matches, i.e. we are comparing $15 \times 6 -2 = 90 - 2 = 88$ lands from unknown bullets against $2 \times 10 \times 6 -2 = 120 - 2 = 118$ lands from known bullets, yielding a total of $10,384$ land-to-land comparisons. Out of these comparisons, there are 172 known matches (KM), while the rest are known non-matches (KNM). 
Ideally, results look like the results in Figure~\ref{fig:hamby-perfect}: Figure~\ref{fig:hamby-perfect}a shows the distribution of the number of maximum consecutive matching striae between land C-3 and all 118 lands from known bullets. Two lands show a high CMS. These correspond to the known matches with C-3, shown in Figures~\ref{fig:hamby-perfect}b and~\ref{fig:hamby-perfect}c. 
 Unfortunately, not all results are as clear cut. 
It might not be reasonable to assume that we can match all lands, but the idea is to try to maximize the number of matches to get an overview of what we might be able to expect from an automated match.

@@

Besides resulting in a probabilistic quantification of matches, random forests also provide an assessment of the importance of each of the features derived from the bullets' 3D topological surface measurements. Figure~\ref{fig:importance} shows an overview of the importance of each variable measured as the mean decrease in the Gini index when the variable in question is included in a tree (for the exact values please refer to Supplement Section~\ref{supp:randomforest}). 
\begin{figure}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=.7\textwidth]{unnamed-chunk-1-1}

@@

%The rows in the table are ordered according to descending importance. 
The variables with the most predictive power are cross-correlation and the overall number of matching extrema, followed by the total depth of joint striations $S$ and total number of non-matches. CMS is found only in sixth place.

Besides including results from known matches against known non-matches, we can increase the number of comparisons in the Hamby study to include all possible land-to-land comparisons. This effectively doubles the number of data points available. Comparisons not previously included in fitting the random forest can also be used as an additional source for assessing error rates.  Results for this and a more detailed discussion can be found in Supplement Section~\ref{supp:extended}.

\section{Discussion}

We present an algorithm which detects the most prominent but least relevant structure of a bullet from a firearms identification perspective, removes these features, and produces residuals which allow for the easy identification of markings. We have generalized this algorithm to align the residuals from two bullets to automatically determine whether they are indistinguishable. A random forest model provides a probabilistic assessment of the strength of a match, along with an ordering of the relevance of features. Matching bullets is clearly not a one-step process, but rather a sequence of data analysis tasks each deserving attention. As there is no scientific standard in place at this point in time, our intent is to explain an approach to addressing these tasks, while documenting all steps and providing all code so other researchers and forensic scientists can reproduce and expand on our findings.

The matching algorithm is sensitive to the parameter choices made. The heights at which signatures are extracted (currently 25$\mu m$ apart) to evaluate stability, as well as the cross-correlation factor (currently 0.95) we set as a minimum threshold do affect the final outcome. Another parameter that must be selected is the amount of smoothing when identifying peaks and grooves (currently, a window of 23.4375$\mu m$ is used, corresponding to a window of 7 values to the left and the right of an observation). We try to lay out in the paper the impact that each of the parameter choices has on the matching performance, but more research and better data are needed to define an optimized scenario.

The Hamby study serves as our evaluation `database'. It consists of only  35 bullets -- this is obviously not a particularly realistic scenario for an automatic matching procedure, but for now we are unaware of other databases containing bullets in the x3p format that we could add to our study.

The feasibility of creating a database of images that could be used to identify guns used in crimes was evaluated in a 2008 report~\citep{nap:2008} by the National Research Council. The committee investigated the scalability of NIBIN (National Integrated Ballistic Information Network), which uses proprietary matching algorithms provided by IBIS. The bottom line of the report was that in spite of the many technical and practical hurdles, solutions to all but one problem could be found. The problem that remained is that statistically, the quality of the matching algorithm (in this case, of breech-face marks and firing pin impressions) could not withstand a hugely increased number of records without overwhelming forensic examiners, who have to examine possible matches suggested by the system. 
The findings of the NRC report on imaging are based on two-dimensional greyscale images, which the committee argued were not reliable enough for distinguishing between fine marks. This finding coincides with the assessment by \citet{dekinder:2004} based on the IBIS Heritage system. A further re-assessment by \citet{deceuster:2015} came to the same conclusions based on the EvoFinder system. 
The NRC report also found that results from 2D images can be improved when matches are based on 3D images. This is consistent with the importance of features found here: out of the top five features (see Figure~\ref{fig:importance}), only the total number of matches and mismatches are available for a match based on 2D features.

By suggesting an automated algorithm that first removes class characteristics, such as the grooves and the curvature of the bullet to reveal the region of the  land, then identifies peaks and valleys on this land, we reduce subjectivity and with it possible sources of bias. In particular, `the concept of counting striations is subjective and based on experience'~\citep{miller:1998}. The steps outlined in this paper could also help explore other important forensic science problems. In particular, more general toolmark examination can benefit from the approach we discuss.

For a fair assessment of the performance of an algorithm, we need transparency. Our matching  algorithm is open: the code is readily available in form of the R package bulletr~\citep{bulletr}, and the code to produce this paper is available at \url{http://www.github.com/erichare/imaging-paper}. To understand whether an automated approach along the lines of the one we propose can accurately identify sets of bullets with undistinguishable markings, it will be necessary to assemble a much larger database that includes a wide range of ammunition types, degrees of damage, gun makes, etc. We are unaware of the existence of any such database. In addition to serving as a realistic testbed for the performance of the automated matching algorithm, such a database would also permit testing the underlying, as of yet untested, assumptions of uniqueness and reproducibility of the markings left by a gun on bullets.
